
```{r include = FALSE}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed, class.output="r-output")
source("./source/r_functions.R")
```



# Introduction {#intro}
(`r format(Sys.time(), '%d %B, %Y, %H:%M')`)

```{css echo=FALSE}
.r-output {
  font-weight: bold;
}
.function {
  background-color: lightgrey;
  padding: 0.5em 0.5em 0.5em 4em;
  border: 1px solid black;
}
```


## Introduction to R

- R is a open-source, freely available _statistical programming language_.
- *R Studio* is the most commonly used interface to R.
- Originally developed at Bell Labs in the 1980s as the _S programming language_.
- Modular, most of the functionality is in code libraries called *packages*. 
- *Literate programming* and *reproducible research* with knitr and markdown.

## The Basics 
- You can think of R as a _really_ fancy calculator, with a _great_ way of handling data.
- The main interaction happens in the *source code window* (by marking code, and pressing Ctrl+R or Apple+R), and on the *R console*.

Addition
```{r}
2+1
```

Subtraction
```{r}
2-1
```

Multiplication
```{r}
2*3
```

Division
```{r}
2/3
```

Exponentiation
```{r}
2^3
```



## Assignment

Values can be stored in **variables**, which have to be named. You might remember the concept from math; the $x$ in $2 = 0.5*x$ is a variable name. 
Variables are created implicitly, by being **assigned** a value. This is done using the R **assignment operator**: `<-`.

For example, let's assign 2 to `var_name`:
```{r}
var_name <- 2
```

Now, let's save the **result** of a **complex expression** ...

```{r}
my_variable <- (11 - 1)^2
my_other_variable <- 10^2
```

... and retrieve it later ...
```{r}
my_variable
```

... or combine with other variables.
```{r}
my_variable + my_other_variable
```


## Evaluation 

- Every **expression** **evaluates to** a value. 

- R displays the value an expression evaluates to for almost every **expression** except assignment.

- The result of the addition is shown:

```{r}
1+1
```

- For the assignment, nothing is shown:

```{r}
x <- 1
```


## Operators 

Variables or expressions can be compared with others.

- The operator **`==`** tests for equality: 

```{r}
my_variable == my_other_variable
```

- The operator **`!=`** tests for inequality:

```{r}
my_variable != my_other_variable
```


- Don't confuse **`==`** with **`=`**. The latter performs assignment (just like **`<-`**):

```{r}
x = 1
```


Variables or expressions can be compared with others.
- What do you think **`<`**, **`>`**, **`<=`**, and **`>=`** test for?

```{r}
my_other_variable == 100
```

```{r}
my_other_variable < 100
```

```{r}
my_other_variable <= 100
```


## Using R Help 

- Let's say you forgot the difference between **`=`**, **`==`**, and **`<-`**.

```{r, eval=FALSE}
help("<-")
```

```{r, eval=F}
help("=")
```

```{r, eval=FALSE}
help("==")
```

```{r, eval=FALSE}
help("glm")
```


## Data Types

In R, values have **types**: 

| Data Type  | Examples|
| -----------:|:-------------------------------:|
| Integer (Numeric): | ..., -3, -2, -1, 0, +1, +2, +3, ... |
| Double (Numeric):  | most rational numbers; e.g., 1.0, 1.5, 20.0, pi|
| Character:  | `"a"`, `"b"`, `"word"`, `"hello dear friend, ..."`|
| Logical:  | `TRUE` or `FALSE` (or: `T` or `F` ) |
| Factor:  | Restricted, user-defined set of values, internally represented numerically (e.g., Gender {'male', 'female', 'other'}) |
| Ordered factor:  | Factor with an ordering (e.g., Starbucks coffee sizes {'venti' > 'grande' > 'tall'}) |


```{r}
my_apples <- 1
my_oranges <- "one"
```

Types matter when comparing: 
```{r, error=T}
my_apples == my_oranges
```


Types also matter for other operations.

```{r}
my_apples <- 1
my_oranges <- "1.1"
```
```{r, error=T}
my_apples + my_oranges
```

```{r, error=T}
my_apples + as.double(my_oranges)
```


## Determining Data Types
When unsure, we can ask R about the type of a variable:

```{r}
my_apples <- 1
my_oranges <- "1.1"
```
```{r, error=T}
typeof(my_apples)
```

```{r, error=T}
typeof(my_oranges)
```

```{r, error=T}
typeof(as.double(my_oranges))
```

## Packages
Most of R's functionality is located in packages contributed by users. In this course, we will use the plotting functionality of `ggplot2`. In chunks like the one below, lines that include a single `#` are the comments. You do not have to run them; they are there to inform you while reading this book. 

```{r, eval=FALSE}
# install ggplot2
install.packages("ggplot2")
# install other packages we will use in this course
install.packages(c("tidyverse", "magrittr", "Hmisc")) 

# load ggplot2
library(ggplot2)
```

Some important functions in R are located in packages. For example, the function `recode` substitutes values by others.
Let's take a look at the function `recode` in the package `dplyr` (`dplyr::recode`, in R lingo). 

```{r, prompt=T}
# Let's define a vector with my favorite fruits
my_favorite_fruit <- c("apples", "pears", "ornges", "mangoes")

# Oh no, I made a typo - it should be 'oranges', not 'ornges'
# Let's replace it, using the function dplyr::recode
dplyr::recode(my_favorite_fruit, 'ornges'='oranges') 
```

As an alternative way of doing the same thing, we can first load the package to gain access to all of its functions.

```{r, prompt=T, message=F}
# Load the dplyr library
library(dplyr)

# Having done this, we can just call this function and all 
# the other functions in  directly, without the `dplyr::`
# prefix.
recode(my_favorite_fruit, 'ornges'='oranges') 

```



## Plotting

One of the great things about R, is that it contains the package `ggplot2`, which produces pretty decent graphics with relatively little effort.

```{r, eval=T}
# Load the ggplot2 package.
library(ggplot2)

# create a vector of the numbers from 1 to 10
x <- 1:10
# now, create a vector of their squares
y <- x^2

# Create a scatterplot of x ~ y, 
# and draw the best fit quadratic function line though them.
# In ggplot2, you can add additional commands by using a single '+' after each command. 
# For ease of reading, I put line breaks for every new command.
# A new line break in an ongoing expression is also shown with a '+' in the beginning of a line. 
# They are not a part of the expression.
ggplot(data=NULL, aes(x, y)) + 
      geom_point() + 
      geom_smooth(method = "lm", formula = y~x + I(x^2) )

```


## Example: Putting it all (and some more) together

Let's try to understand the Starbucks pricing system.

Here is how we get our data into R.

```{r, error=T}

# store size labels in a vector
label <- c('tall', 'grande', 'venti')
# make it an ordered vector
label <- ordered(label, c('tall', 'grande', 'venti'))

# store sizes in ounces in a vector
size_oz <- c(12, 16, 20)

# store cafe latte prices in a vector; prices probably not up to date
price_tl <- c(8.50, 9.75, 11.00) 

df_starbucks_coffee <- data.frame(label, size_oz, price_tl)
df_starbucks_coffee
```



```{r, error=T}

library(ggplot2)

# plot cup size by size category, use a bar plot;
ggplot(df_starbucks_coffee, aes(label, size_oz)) + 
  geom_bar(stat = "identity")

```

```{r, error=T}

# plot cup size by price, use a scatterplot;  draw a line through the points
ggplot(df_starbucks_coffee, aes(price_tl, size_oz)) + 
  geom_point()  + 
  geom_smooth(method = "lm")

```

### How much would other coffee sizes cost if Starbucks introduced them?

```{r, error=T}

# plot cup size by price, use a scatterplot;  draw a line through the points
ggplot(df_starbucks_coffee, aes(price_tl, size_oz)) + 
  geom_point()  + 
  geom_smooth(method = "lm", fullrange=TRUE) + 
  scale_x_continuous(limits = c(3, 15))  + 
  scale_y_continuous(limits = c(-3, 25))

```

- Does this plot even make sense? Why not?

...

- The above plot isn't the most intuitive way to visualize the price-coffee size relationship, because we tend to interpret such plots in terms of 'a change of a certain number of units on the x-axis *leads* to a change of a certain number of units on the y-axis'.
- In the above case, this means that *'if I choose to pay approximately 5 lira, I'll get 0 ounces of coffee'*, or *'if I choose to pay 4 lira, I'll get -2.5 ounces of coffee'*. This is of course nonsensical. 

### How much would other coffee sizes cost if Starbucks introduced them? A second attempt

- Flipping the coordinates, with coffee size on the x-axis, and price on the y-axis makes it *easier* to see that the only useful part of the plot is where the coffee size is larger than $0$, since you'd never ask yourself the question *'How much would I need to pay, in order to _give_ Starbucks 1 ounce of coffee?'*.

```{r, error=T}

# plot cup size by price, use a scatterplot;  draw a line through the points
ggplot(df_starbucks_coffee, aes(size_oz, price_tl)) + 
  geom_point()  + 
  geom_smooth(method = "lm", fullrange=TRUE) + 
  scale_x_continuous(limits = c(-1, 25))  + 
  scale_y_continuous(limits = c(0, 15))

```

## Operators and functions in this section

### Operators
`r function_definition("x + y", "Addition")`
<br>
`r function_definition("x - y", "Subtraction")`
<br>
`r function_definition("x * y", "Multiplication")`
<br>
`r function_definition("x / y", "Division")`
<br>
`r function_definition("x^y", "Exponentiation")`
<br>
`r function_definition("x <- y", "Assignment")`
<br>
`r function_definition("==", "Test for equality. **Don't confuse with a single =, which is an assignment operator (and also always returns TRUE).**")`
<br>
`r function_definition("!=", "Test for inequality")`
<br>
`r function_definition("<", "Test, smaller than")`
<br>
`r function_definition(">", "Test, greater than")`
<br>
`r function_definition("<=", "Test, smaller than or equal to")`
<br>
`r function_definition(">=", "Test, greater than or equal to")`
<br>

### Functions

```{r results='asis', echo=F}  
cat( function_definition("install.packages(package_name)", 
                         "Installs one or several package(s). 
                         The argument `package_name` can either be a character (`install.packages('dplyr')`) like or a character vector (`install.packages(c('dplyr','ggplot2'))`)."))
```
<br>
```{r results='asis', echo=F}  
cat( function_definition("library(package_name)", 
                         "Loads a package called `package_name`."))
```
<br>
`r function_definition("typeof(x)", "Determines the type of a variable/vector.")`
<br>
`r function_definition("as.double(x)", "Converts a variable/vector to type **double**.")`
<br>
